Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-24 paper2.qmd
section:theory; process:write

<!-- 3. But what critical issues remain in CJ Research?, what is the organization of the study? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved. This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates. The first issue concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

The study has five main sections. It begins with a brief overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues. In @sec-theory, the study introduces a theoretical model for CJ that builds upon Thurstone's theory, along with its statistical translation, designed to address both issues simultaneously. Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion presents the conclusions.

@@

This section examines the two critical issues in CJ's theory that can compromise the reliability and validity of CJ's trait estimates. In @sec-theory-issue1, the study examines the first issue: the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. In @sec-theory-issue2, the study explores the second issue: the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

# Updating the theoretical and statistical model {#sec-theory}

The objective of this section is twofold. First, it aims to refine CJ's theoretical model using the *structural approach* to causal inference. This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory and key assessment design features relevant to CJ experiments. Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data.

@@

## The theoretical model {#sec-theory-theoretical}

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system. Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference. They achieve this by helping researchers to determine which statistical models can identify an estimand, assuming the depicted causal structure is correct [@McElreath_2020]. *Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4].

<!-- 2. The SCMs -->
SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

<!-- 3. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables. The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 4. The benefits of SCMs and DAGs -->
SCMs and DAGs support causal inference through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35].

<!-- 5. what are we doing with them? -->
Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments. This process leads to the development of *the sample-comparison model* in @sec-theory-theoretical_SC.

### The population model {#sec-theory-theoretical_P}

@@

![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%}

![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%}

![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%}

@@

![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%}

![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%}

![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%}

![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%}

@@

### The sample-comparison model {#sec-theory-theoretical_SC}

@@

![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%}

@@

![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%}

@@

## From theory to statistics {#sec-theory-statistics}

@@

<!-- 2. What are the benefits of Bayesian inference?-->
Bayesian inference procedures offer three key advantages. First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999]. Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds. This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].

@@

<!-- 2. What do they do next? -->
Researchers would then proceed to identify the estimands. *Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. An *estimator* is a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand, and they are derived through *estimation*, referring to the process of integrating data with an estimator [@Everitt_et_al_2010]. The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020] in @fig-IEflow provides a visual representation of the process of transitioning from estimands to estimates.

@@

<!-- 3. But, what do we need to produce a good estimate? -->
While numerous estimators can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006]. If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively. The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.

@@

<!-- 4. But we want these estimands to be causal -->
However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations. In the example, researchers can interpret the associational estimand in @eq-group_diff and the associational estimate represented by the Z-statistic as causal. This interpretation is valid as long as the data satisfy the assumptions of the Z-test and are collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. This recognition stems from their ability to enable researchers interpret associational estimates as causal. They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010]. Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].

@@

<!-- 1. The benefits of SCMs and DAGs -->
The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference. They achieve this by helping researchers to determine which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020]. @fig-IEflow shows the role of theoretical models in the inference process.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved. This study addresses two such issues that can compromise the reliability and validity of CJ's trait estimates. The first concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

<!-- 4. what is the organization of the study? -->
The study is organized into six main sections. It begins with an overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues. In @sec-theoretical, the study develops a theoretical model for CJ that simultaneously addresses these issues. The model integrates Thurstone's core principles with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a statistical model that supports the analysis of pairwise comparison data. Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion summarizes the findings.

@@

This section explores the two critical issues in CJ's theory that have the potential to compromise the reliability and validity of CJ's trait estimates. @sec-theory-issue1 inspects the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

# Updating the theoretical model {#sec-theoretical}

<!-- This study examines two such problems that can compromise the reliability and validity of CJ's trait estimates. -->

<!-- This section aims to refine CJ's theoretical model using the *structural approach* to causal inference with the purpose of addressing the two.  -->

<!-- This approach, as outlined in @sec-theory-theoretical, defines a model that integrates the core principles of Thurstone's theory, as well as other key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and the number of comparisons.  -->

@@

<!-- 1. The structural approach to causal inference -->
<!-- The *structural approach* to causal inference provides a formal framework to identify causes and estimate their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system. Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference. They achieve this by helping researchers to determine which statistical models can identify an estimand, assuming the depicted causal structure is correct [@McElreath_2020]. *Identification* determines whether an estimator can accurately compute an estimand based on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. -->

<!-- 2. The SCMs -->
<!-- SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence. -->

<!-- 3. The DAGs -->
<!-- Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables. The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020]. -->

<!-- 4. The benefits of SCMs and DAGs -->
<!-- SCMs and DAGs support causal inference through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35].  -->

<!-- 5. what are we doing with them? -->
<!-- Therefore, this section leverages SCMs and DAGs to represent the conceptual population of CJ experiments, referred to as *the population model* in @sec-theory-theoretical_P, and later integrates various assessment design features relevant to CJ experiments. This process leads to the development of *the sample-comparison model* in @sec-theory-theoretical_SC. -->

## The population model {#sec-theory-theoretical_P}

@@

<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->

<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->

<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->

@@

<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->

<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->

<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->

<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->

@@

## The sample-comparison model {#sec-theory-theoretical_SC}

@@

<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->

@@

<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->

@@

# From theory to statistics {#sec-statistical}

<!-- Second, @sec-theory-statistics uses *Bayesian inference methods* to convert these theoretical and practical elements into a statistical model that supports the analysis of paired comparison data. -->

@@

<!-- 2. What are the benefits of Bayesian inference?-->
<!-- Bayesian inference procedures offer three key advantages. First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999]. Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds. This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014]. -->

@@

<!-- 2. What do they do next? -->
Ideally, researchers then proceed to identify the estimands. *Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Identification is crucial because if an estimator cannot reliably compute an estimand using "infinite" and error-free data, it is impossible to gain meaningful insights about that estimand using finite data [@Schuessler_et_al_2023, pp. 5]. An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand derived through the process of *estimation*, which integrates data with an estimator [@Everitt_et_al_2010]. The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], shown in @fig-IEflow, visually represents the transition from estimands to estimates.

@@

<!-- 3. But, what do we need to produce a good estimate? -->
Consequently, although many estimators can approximate an estimand, researchers prioritize those with desirable identification properties to ensure accurate and reliable estimates. For instance, the Z-test is a widely used estimator for comparing group proportions, providing accurate estimates when its underlying assumptions are satisfied [@Kanji_2006]. Moreover, researchers can interpret the Z-test estimates as causal, provided they collect the data through a randomized experiment.

<!-- If both assumptions are met, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively. The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.  -->

@@

<!-- However, researchers often seek to uncover the mechanisms underlying specific data and establish causal relationships rather than simply estimate associations. -->

<!-- 4. Why does causality stem from randomize experiments? -->
Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. This recognition stems from their ability to enable researchers interpret associational estimates as causal. They achieve this by ensuring data, and by extension an estimator, satisfies several key identification properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010]. Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].

@@

<!-- 1. The benefits of SCMs and DAGs -->
The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. Essentially, these tools serve as *conceptual (theoretical) models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. Specifically, they guide researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020], thus enabling valid causal inference. @fig-IEflow shows the role of theoretical models in the inference process.
